---
title: "Lab 2 - Data wrangling"
author: "Deandra Rasheesa Maheswari"
date: "February 4, 2026"
format: 
  html: default # for speed
  pdf: default # for final output
---

```{r}
#| eval: true
#| message: false

library(tidyverse)
```

# Questions

## Part 1
```{r}
?midwest
```
### Question 1
```{r}
midwest |>
  count(state, name = "n_counties") |>
  arrange(desc(n_counties))
```
State with the highest number of counties is Illinois with 102 counties, while state with the lowest number of counties is Wisconsin with 72 counties.
\newpage

### Question 2
```{r}
midwest |>
  count(county, name = "n_states") |>
  filter(n_states == 5) |>
  arrange(desc(n_states))
```
Counties with the names Crawford, Jackson and Monroe appear in all five states 
in the dataset.
\newpage

### Question 3
```{r}
midwest |>
  filter(popdensity > 25000) |>
  select(county, state, popdensity, poptotal, area) |>
  arrange(desc(popdensity))
```

```{r}
midwest |>
  filter(popdensity == max(popdensity)) |>
  select(county, state, popdensity, poptotal, area)
```
There are nine counties with population density above 25000 as can be seen on table above in part (a) and county with highest population density is Cook County in Illinois with 88018.4 person per unit area.
\newpage

### Question 4
```{r}
midwest |>
  summarize(
    median_density = median(popdensity),
    q1_density = quantile(popdensity, 0.25),
    q3_density = quantile(popdensity, 0.75)
  )
```
The distribution of population density of counties is unimodal and extremely right-skewed. A typical Midwestern county has a population density of 1156.208 people per unit area. The middle 50% of the counties have population densities between 622.4074 and 2330 people per unit area.
\newpage

### Question 5
```{r}
midwest |>
  mutate(metro = if_else(inmetro == 1, "Yes", "No")) |>
  group_by(state) |>
  summarize(propotion_urban = mean(metro == "Yes"))
```
Ohio has the highest proportion of urban (metro) counties with about 0.45 (45%) of its counties classified as metro areas. On the other hand, Illinois and Wisconsin have the lowest proportion of urban counties among the states in this dataset.
\newpage

### Question 6

\newpage

### Question 7

\newpage

### Question 8

\newpage

## Part 2

### Question 9
